Fine scale determinants of wood and leaf strategies in a low-elevation rainforest of India: vertical stratification reflects guild-dependent dynamics

### 2016-08-12 inst/doc/draft.Rmd
v0.1.9034 Corrected strata beta results

Understanding the determinants of species occurrence and co-occurrence is a long standing issue in ecology [@diamond_ecology_1975]. Tropical rainforests have especially fascinated standing issue in ecology [@connell_diversity_1978]. Two concurrent theories have been proposed to explain how species coexistence can be maintained in ecological communities. Niche theory states that the niche differences of species determine wether they can establish and coexist in communities [@weiher_assembly_1995; @lortie_rethinking_2004], while neutral theory shows that species coexistence is possible even when species are functionaly equivalent, under the sole influence of stochasticity in life, death, reproduction and dispersal [@hubbell_unified_2001]. Both theory can provide realistic expectations of taxonomic diversity [@chave_comparing_2002], and some recent approaches have advocated investigating patterns of functional [@westoby_land-plant_2006; @mcgill_rebuilding_2006] and phylogenetic diversity [@mouquet_ecophylogenetics:_2012] to trace the signature of ecological processes related to specific biological attributes and adaptations. 
These approaches have been applied to resolve the conundrum of high tree diversity in tropical forests [@kraft_functional_2010; @baraloto_using_2012; @hardy_phylogenetic_2012], but studies on these ecosystems still remain scarce. We here investigate taxonomic, functional and phylogenetic diversity in a low elevation rainforest of the Western Ghats biodiversity hotspot [@myers_biodiversity_2000], where no study has addressed the relative influence of neutral and niche-based dynamics on local forest composition.

@@

### Taxonomic, functional and phylogenetic variation among communities

To investigate taxonomic, functional, and phylogenetic variation among communities we calculated $I_{st}$, for taxonomic composition, $U_{st}$ and $\tau _{st}$, for functional composition in abundance and presence-absence respectively, and $\beta _{st}$ and $\pi _{st}$, for phylogenetic composition in abundance and presence-absence respectively [@hardy_characterizing_2007]. $I_{st}$ measures species turnover accounting for local abundance. $U_{st}$ and $\beta _{st}$ integrate only phylogenetic and functional turnover, respectively, weighting species by their local abundances; whereas $\tau _{st}$ and $\pi _{st}$ express phylogenetic and functional turnover, respectively, without accounting for species local abundances.

@@

We used null models to test if observed correlation with Mantel statistic differed significantly from expectations under null hypothesis. $I_{st}$ correlation to environment and space was tested with 999 repetions of the model nul described by @myers_beta-diversity_2013 which randomizes individuals among subplots while preserving the regional abundance of species, and the number of individuals per community. $U st$ and $\tau st$ were tested by the nul model NM1 which randomizes traits among species. Finally $\beta st$ and $\pi st$ were tested against the nul model *"1s"* suggested by @hardy_testing_2008 which randomizes phylogenetic tree tips.

@@

Table 3: Summary of linear models of community-level trait values. _The results are shown for either CWM values weighted either by species relative abundance (AB) or in presence-absence (PA) after stepwise selection of the predictors including slope, curvature, wetness, southwesterness (SW), canopy height (Canopy) and basal area (BA). Global and partial F-statistics of final models were compared to statistics of 999 null linear regressions built with null model NM1, which randomizes traits among species. See table 1 for abbreviations. The number of asterisk indicates the significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001)._

@@

Traits convergence/divergence was measured with community weighted variances (CWV). Overall, observed CWVs did not differ from null expectations for all traits. We could found significant relationships between standardized effect size (SES) of CWVs and environmental variables (see table 4). Convergence of LES and LDMC and divergence of LA were found under south-west orientation. LA and LDMC values also diverged in communities with increasing curvature. Observed CVNNDs did not reveal niche differentiation among communities for all traits. But we also found significant relationships of their standardized effect size (SES) of CVNNDs with environmental variables (see table 4). niche differentiation of LA increased in lower curvature and wetness contexts, while differentiation of WD decreased in southwest orientation.

Table 4: Summary of linear models of traits standardized effect size (SES) of CWVs and CVNNDs. _Linear regression results for each trait SES of CWVs and CVNNDs after stepwise selection of the predictors oncluding slope, curvature, wetness southwesterness (SW), canopy height (Canopy) and basal area (BA). See table 1 for abbreviations. The number of asterisk indicates the significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001)._

@@

```{r CWV & CVNND table}
res <- list(CWV = CWV.ses.res, CVNND = CVNND.ses.res)
res <- mapply(function(s, name){
  s <- cbind(c(name, rep(' ', length(s[,1])-1)), s)
  names(s)[1:2] <- c('T', 'S')
  return(s)
}, s = res, name = as.list(c('CWV', 'CVNND')), SIMPLIFY = F)
res <- do.call('rbind', res)
kable(bold(res), row.names = F)
rm(CWV.ses.res, CVNND.ses.res, CVNND.pval.res, res)
```

## Taxonomic, functional and phylogenetic varitions among communities

Taxonomic variation among communities was correlated significantly to environment variation (Mantel test, $r_{I_{st},E} = 0.23$ with $p-value < 0.001$), but not with remaining spatial dependency in community composition (Mantel test of space knowing environment, $r_{I_{st},S|E} = 0.01$ with $p-value > 0.1$).

Overall, the global functional variation among communities was neither correlated to environment nor to spactial distance knowing the environment effect (see table 5). However, the variation of WD, in abundance as well as in presence absence, was positively correlated to environmental variation and is negatively correlated in presence absence to the spatial structure knowing environment spatial effect.

```{r Ist}
library(spacodiR)

# R,L,Q wihtout strata
R <- com@data[c('Slope', 'Curvature', 'Wetness', 'SW', 'Canopy', 'BA')]
row.names(R) <- com@data$id
L <- data.frame(tapply(rep(1, length(Trees[,1])), list(Trees$com, Trees$SpCode), sum))
L[is.na(L)] <- 0
Q <- PFT_sp[c('LES', 'SLA', 'WES', 'WD', 'LDMC', 'LA', 'LT')]
row.names(Q) <- PFT_sp$SpCode
sel <- intersect(row.names(Q), names(L))
Q <- Q[sel,]
L <- L[sel]
R <- R[row.names(L),]
L <- list(abundance = L, `presence-absence` = L)
L$`presence-absence`[L$`presence-absence` > 1]  <- 1
rm(com, Trees)
invisible(gc())

# Environmental matrices
XY <- data.frame(as(quadrats(), 'SpatialPoints')@coords)
row.names(XY) <- paste0('C',row.names(XY))
XY <- XY[row.names(L$abundance),]
Ms <- dist(XY)
XY <- pca.env$scores[,1:2]
row.names(XY) <- paste0('C',row.names(XY))
XY <- XY[row.names(L$abundance),]
Me <- dist(XY)
rm(XY)

# Observed Ist
spaflo <- spacodi.calc(sp.plot = t(L$abundance), pairwise = T)$'pairwise.Ist'
## Mantel values
spaflo.mantel <- list(`E` = vegan::mantel(spaflo, Me, permutations = 0)$statistic, 
                      `S | E`= vegan::mantel.partial(spaflo, Ms, Me, permutations = 0)$statistic)

# Simulated mantel
nL <- vegan::permatfull(L$abundance, fixedmar = 'both', mtype = 'count', times = n, shuffle = 'both')
cl <- makeCluster(cores)
clusterExport(cl, list('nL', 'spacodi.calc', 'Me', 'Ms'))
nspaflo <- lapply(nL$perm, function(l){spacodi.calc(sp.plot = t(l), pairwise = T)$'pairwise.Ist'})
nspaflo.mantel <- lapply(nspaflo, function(ns){
  list(`E` = vegan::mantel(ns, Me, permutations = 0)$statistic, 
       `S | E`= vegan::mantel.partial(ns, Ms, Me, permutations = 0)$statistic)})
stopCluster(cl)
rm(cl, nL) ; invisible(gc())
## Non null mantels
nspaflo.mantel <- lapply(nspaflo.mantel, `[`, names(nspaflo.mantel[[1]]))
nspaflo.mantel <- apply(do.call(rbind, nspaflo.mantel), 2, as.list)
nspaflo.mantel <- lapply(nspaflo.mantel, unlist)
spaflo.res <- mapply(function(s, ns){
    paste(format(s, scientific = T, digits = 3),
          Uppangala::stars(min(1-rank(c(s,ns))[1]/(n+1),rank(c(s,ns))[1]/(n+1)), ns = '   '))
}, s = spaflo.mantel, ns = nspaflo.mantel)
spaflo.res <- data.frame(spaflo.res)
names(spaflo.res) <- c('$I_{st}$')

# Result
# kable(bold(spaflo.res))
rm(mtest, spaflo.mantel, nspaflo, nspaflo.mantel, spaflo.res)
```

Table 5: Summary of Mantel tests between functional composition, environment and spatial distances between communities. _$U st$ and $\tau st$ metric proposed by @hardy_characterizing_2007 were used to measure functional composition variation among communities. Observed values were compared to null values obtained with model NM1 randomizing traits among species (permuting raw in `Trait*Species` matrix). E is the Mantel test of this metric according to the environmental distance derived from communtiy scores in the environmental PCA (two first axes). S|E is the partial Mantel test of this metric according to spatial distance conditionally to environment variation. The number of asterisk indicates significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001)._

```{r Ust & TAUst}
# nQ without strata
n = 999
nQ <- rep(list(Q), n)
nQ <- lapply(nQ, function(x){row.names(x) <- sample(row.names(x)) ; return(x)})
nQ <- lapply(nQ, function(x){x[names(L$abundance),]})

# Observed Hardys
spafun <- lapply(as.list(1:length(Q)), function(i){
  if (length(which(is.na(Q[i]))) > 0) {
    sel <- intersect(names(L$abundance), row.names(Q[-which(is.na(Q[i])),]))
    return(spacodi.calc(sp.plot = t(L$abundance[sel]), sp.traits = data.frame(Q[sel,i], row.names = sel), pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
  } else {
    return(spacodi.calc(sp.plot = t(L$abundance), sp.traits = Q, pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
  }
})
names(spafun) <- names(Q)
## Mantel values
spafun.mantel <- lapply(spafun, function(t){list(`E` = lapply(t, function(M){vegan::mantel(M, Me, permutations = 0)$statistic}), `S | E` = lapply(t, function(M){vegan::mantel.partial(M, Ms, Me, permutations = 0)$statistic}))})

# Simulated mantel
cl <- makeCluster(cores)
clusterExport(cl, list('nQ', 'L', 'Me', 'Ms', 'spacodi.calc'))
nspafun.mantel <- clusterApply(cl, nQ, function(q){
  s <- lapply(as.list(1:length(q)), function(i){
    if (length(which(is.na(q[i]))) > 0) {
      sel <- intersect(names(L$abundance), row.names(q[-which(is.na(q[i])),]))
      return(spacodi.calc(sp.plot = t(L$abundance[sel]), sp.traits = data.frame(q[sel,i], row.names = sel), pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
    } else {
      return(spacodi.calc(sp.plot = t(L$abundance), sp.traits = q, pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
    }
  })
  names(s) <- names(q)
  res <- list(
    `E` = lapply(s, function(t){
      lapply(t, function(M){
        try(vegan::mantel(M, Me, permutations = 0)$statistic)
        })
      }),
    `S | E` = lapply(s, function(t){
      lapply(t, function(M){
        try(vegan::mantel.partial(M, Ms, Me, permutations = 0)$statistic)
        })
      })
    )
  res <- lapply(res, function(r){names(r) <- names(q) ; return(r)})
  res <- lapply(res, `[`, names(res[[1]]))
  res <- apply(do.call(rbind, res), 2, as.list)
  return(res)
})
stopCluster(cl)
rm(cl, nQ) ; invisible(gc())

## Non null mantels
nspafun.mantel <- lapply(nspafun.mantel, `[`, names(nspafun.mantel[[1]]))
nspafun.mantel <- apply(do.call(rbind, nspafun.mantel), 2, as.list)
nspafun.mantel <- lapply(nspafun.mantel, function(s){lapply(s, `[`, names(s[[1]]))})
nspafun.mantel <- lapply(nspafun.mantel, function(s){apply(do.call(rbind, s), 2, as.list)})
nspafun.mantel <- lapply(nspafun.mantel, function(s){lapply(s, function(m){lapply(m, `[`, names(m[[1]]))})})
nspafun.mantel <- lapply(nspafun.mantel, function(s){lapply(s, function(m){apply(do.call(rbind, m), 2, as.list)})})
nspafun.mantel <- lapply(nspafun.mantel, function(s){lapply(s, function(x){lapply(x, function(x){as.numeric(unlist(x))})})})
spafun.res <- data.frame(mapply(function(x,y){
  mapply(function(r,s){
    mapply(function(p,q){
      paste(round(p, 3), Uppangala::stars(min(rank(c(p,q))[1]/(n+1), 1-rank(c(p,q))[1]/(n+1)), ns = '   '))
    }, p = r, q = s, SIMPLIFY = F)
  }, r = x, s= y)
}, x = spafun.mantel, y = nspafun.mantel))
spafun.res <- cbind(data.frame('Mantel statistic' = c('E', ' ', 'S|E', ' '), 'Metric' = rep(c('$U_{st}$', '$\\tau _{st}$'), 2)), spafun.res)

# Result
kable(bold(spafun.res))
rm(spafun.mantel, nspafun.mantel, spafun.res)
```

```{r BETAst $ PIst}
# Observed Hardys
phy <- tree
phy$tip.label <- Species$SpCode[match(gsub('_', ' ',phy$tip.label), Species$LatinName)]
tips <- c(setdiff(phy$tip.label, names(L$abundance)), setdiff(names(L$abundance), phy$tip.label))
phy <- drop.tip(phy, tips)
phy$node.label <- NULL
spaphy <- spacodi.calc(sp.plot = t(L$abundance), phy = phy, pairwise = T)[c('pairwise.Bst', 'pairwise.PIst')]
## Mantel values
spaphy.mantel <- list(`E` = lapply(spaphy, function(m){vegan::mantel(m, Me, permutations = 0)$statistic}), 
                      `S | E`= lapply(spaphy, function(m){vegan::mantel.partial(m, Ms, Me, permutations = 0)$statistic}))

# Simulated mantel
nphy <- rep(list(phy), 999)
cl <- makeCluster(cores)
clusterExport(cl, list('nphy', 'L', 'spacodi.calc', 'Ntip', 'resamp.phy', 'branching.times'))
nphy <- clusterApply(cl, nphy, resamp.phy)
stopCluster(cl)
nspaphy <- lapply(nphy, function(p){spacodi.calc(sp.plot = t(L$abundance), phy = p, pairwise = T)[c('pairwise.Bst', 'pairwise.PIst')]})
rm(cl, nphy) ; invisible(gc())
nspaphy.mantel <- list(`E` =lapply(nspaphy, function(x){lapply(x, function(M){vegan::mantel(M, Me, permutations = 0)$statistic})}), `S | E` = lapply(nspaphy, function(x){lapply(x, function(M){vegan::mantel.partial(M, Ms, Me, permutations = 0)$statistic})}))
## Non null mantels
nspaphy.mantel <- lapply(nspaphy.mantel, function(s){lapply(s, `[`, names(s[[1]]))})
nspaphy.mantel <- lapply(nspaphy.mantel, function(s){apply(do.call(rbind, s), 2, as.list)})
nspaphy.mantel <- lapply(nspaphy.mantel, function(s){lapply(s, unlist)})
spaphy.res <- mapply(function(s, ns){
  mapply(function(x,y){
    paste(format(x, scientific = T, digits = 3), 
          Uppangala::stars(min(1-rank(c(x,y))[1]/(n+1),rank(c(x,y))[1]/(n+1)), ns = '   '))
  }, x = s, y = ns)
}, s = spaphy.mantel, ns = nspaphy.mantel)
row.names(spaphy.res) <- c('$\\beta_{st}$', '$\\pi_{st}$')

# Result
# kable(bold(spaphy.res))
rm(mtest, spaphy.mantel, nspaphy, nspaphy.mantel, spaphy.res)
```

@@

>>>>

Understanding the determinants of species occurrence and co-occurrence is a long standing issue in ecology [@diamond_ecology_1975]. Tropical rainforests have especially fascinated ecologists due to their outstanding diversity [@connell_diversity_1978]. Many theories have been proposed to explain how species coexistence can be maintained in ecological communities . Niche theory states that the niche differences of species determine wether they can establish and coexist in communities [@weiher_assembly_1995; @lortie_rethinking_2004], while neutral theory shows that species coexistence is possible even when species are functionaly equivalent, under the sole influence of stochasticity in life, death, reproduction and dispersal [@hubbell_unified_2001]. Both theory can provide realistic expectations of taxonomic diversity [@chave_comparing_2002], and some recent approaches have advocated investigating patterns of functional [@westoby_land-plant_2006; @mcgill_rebuilding_2006] and phylogenetic diversity [@mouquet_ecophylogenetics:_2012] to trace the signature of ecological processes related to specific biological attributes and adaptations. 
These approaches have been applied to resolve the conundrum of high tree diversity in tropical forests [@kraft_functional_2010; @baraloto_using_2012; @hardy_phylogenetic_2012], but studies on these ecosystems still remain scarce. We here investigate taxonomic, functional and phylogenetic diversity in a low elevation rainforest of the Western Ghats biodiversity hotspot [@myers_biodiversity_2000], where no study has addressed the relative influence of neutral and niche-based dynamics on local forest composition.

@@

### Taxonomic and functional variation among communities

To investigate taxonomic and functional variation among communities we calculated $I_{st}$, for taxonomic composition and $U_{st}$ and $\tau _{st}$, for functional composition in abundance and presence-absence respectively [@hardy_characterizing_2007]. $I_{st}$ measures species turnover accounting for local abundance. $U_{st}$ integrate only functional turnover, weighting species by their local abundances; whereas $\tau _{st}$ express functional turnover, without accounting for species local abundances.

@@

We used null models to test if observed correlation with Mantel statistic differed significantly from expectations under null hypothesis. $I_{st}$ correlation to environment and space was tested with 999 repetions of the model nul described by @myers_beta-diversity_2013 which randomizes individuals among subplots while preserving the regional abundance of species, and the number of individuals per community. $U st$ and $\tau st$ were tested by the nul model NM1 which randomizes traits among species.

@@

Table 3: Summary of linear models of traits standardized effect size (SES) of CWMs. _Linear regression results for each trait SES of CWMs, for canopy (C) and understorey (U) guilds, in abundance (AB) and presence-absence (PA), after stepwise selection of the predictors including slope, curvature, wetness southwesterness (SW), canopy height (Canopy) and basal area (BA). See table 1 for abbreviations. The number of asterisk indicates the significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001)._

@@

Traits convergence/divergence was measured with community weighted variances (CWV). We could found significant relationships between standardized effect size (SES) of CWVs and environmental variables (see table 4). Convergence of WES and LDMC in the canopy guild  and convergence of LA in the undersotrey guild were found under concave areas. LDMC of the canopy guild converged also with canopy height. LES of understorey guild converged in slopes and SLA of understorey guild diverged with increasing BA. We also found significant relationships of their standardized effect size (SES) of CVNNDs with environmental variables (see table 4). Niche differentiation of WES and LDMC in canopy guild increased in lower basal area context. Differentiation of canopy guild LA increased in convex and high canopy contexts, whereas LT of the canopy guild differentiated in concave context. Niche differentiation of LES in understorey guild increased in concave areas, while differentiation of LT of the understorey guild increased in wet context with a high basal area.

Table 4: Summary of linear models of traits standardized effect size (SES) of CWVs and CVNNDs. _Linear regression results for each trait SES of CWVs and CVNNDs, for canopy (C) and understorey (U) guilds, after stepwise selection of the predictors including slope, curvature, wetness southwesterness (SW), canopy height (Canopy) and basal area (BA). See table 1 for abbreviations. The number of asterisk indicates the significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001)._

@@

```{r CWV & CVNND table}
res <- list(CWV = CWV.ses.res, CVNND = CVNND.ses.res)
res <- mapply(function(s, name){
  s <- cbind(c(name, rep(' ', length(s[,1])-1)), s)
  names(s)[1:2] <- c('T', 'S')
  return(s)
}, s = res, name = as.list(c('CWV', 'CVNND')), SIMPLIFY = F)
res <- do.call('rbind', res)
kable(bold(res), row.names = F)
rm(CWV.ses.res, CVNND.ses.res, CVNND.pval.res, sel, res)
```

## Taxonomic and functional varitions among communities

Taxonomic variation among communities was correlated significantly to environment variation for both canopy and understorey guilds (Mantel test for canopy and understorey guild, $r_{I_{st},E,C} = 0.22$ and $r_{I_{st},E,U} = 0.23$, respectively, with $p-value < 0.01$), but not with remaining spatial dependency in community composition for canopy guild (Mantel test of space knowing environment, $r_{I_{st},S|E,C} = 0.01$ with $p-value > 0.1$), whereas understorey guild taaxonomic variation was correlated to remaining spatial dependency in community composition (Mantel test of space knowing environment, $r_{I_{st},S|E,U} = 0.11$ with $p-value < 0.01$).

```{r Ist, eval=FALSE, include=FALSE}
library(spacodiR)

# Environmental matrices
XY <- data.frame(as(quadrats(c(30,30)), 'SpatialPoints')@coords)
row.names(XY) <- paste0('C',row.names(XY))
XY <- lapply(L, function(s){
  XY[row.names(s$abundance),]
})
Ms <- lapply(XY, function(s){
  dist(s)
})
XY <- pca.env$scores[,1:2]
row.names(XY) <- paste0('C',row.names(XY))
XY <- lapply(L, function(s){
  XY[row.names(s$abundance),]
})
Me <- lapply(XY, function(s){
  dist(s)
})
rm(XY)

# Observed Ist
spaflo <- lapply(L, function(s){
  spacodi.calc(sp.plot = t(s$abundance), pairwise = T)$'pairwise.Ist'
})
## Mantel values
spaflo.mantel <- mapply(function(s, Mes, Mss){
  list(
    `E` = vegan::mantel(s, Mes, permutations = 0)$statistic, 
    `S | E`= vegan::mantel.partial(s, Mss, Mes, permutations = 0)$statistic
  )
}, s = spaflo, Mes = Me, Mss = Ms, SIMPLIFY = F)

# Simulated mantel
nL <- lapply(L, function(s){
  vegan::permatfull(s$abundance, fixedmar = 'both', mtype = 'count', times = n, shuffle = 'both')
})
nspaflo <- lapply(nL, function(s){
  lapply(s$perm, function(l){
    spacodi.calc(sp.plot = t(l), pairwise = T)$'pairwise.Ist'
  })
})
nspaflo.mantel <- mapply(function(s, Mes, Mss){
  lapply(s, function(x){
    try(list(
      `E` = vegan::mantel(x, Mes, permutations = 0)$statistic, 
      `S | E`= vegan::mantel.partial(x, Mss, Mes, permutations = 0)$statistic
    ))
  })
}, s = nspaflo, Mes = Me, Mss = Ms, SIMPLIFY = F)
sel <- which(unlist(lapply(nspaflo.mantel$U, inherits, 'try-error')))
nspaflo.mantel$U <- nspaflo.mantel$U[-sel]
## Non null mantels
nspaflo.mantel <- lapply(nspaflo.mantel, function(s){
  lapply(s, `[`, names(s[[1]]))
  apply(do.call(rbind, s), 2, as.list)
})
nspaflo.mantel <- lapply(nspaflo.mantel, function(s){
  lapply(s, unlist)
})
spaflo.res <- mapply(function(s, ns){
  mapply(function(x, nx){
    paste(format(x, scientific = T, digits = 3),
          Uppangala::stars(min(1-rank(c(x,nx))[1]/(n+1),rank(c(x,nx))[1]/(n+1)), ns = '   '))
  }, x =s, nx = ns)
}, s = spaflo.mantel, ns = nspaflo.mantel)

# Result
kable(bold(spaflo.res))
rm(spaflo.mantel, nspaflo, nspaflo.mantel, spaflo.res)
```

Most of trait functional variation among communities was neither correlated to environment nor to spactial distance knowing the environment effect (see table 5). However, the variation of WD, in abundance as well as in presence absence, was positively correlated to environmental variation and is negatively correlated in presence absence to the spatial structure knowing environment spatial effect.

Table 5: Summary of Mantel tests between functional composition, environment and spatial distances between communities. _$U st$ and $\tau st$ metric proposed by @hardy_characterizing_2007 were used to measure functional composition variation among communities, for canopy (C) and understorey (U) guilds. Observed values were compared to null values obtained with model NM1 randomizing traits among species (permuting raw in `Trait*Species` matrix). E is the Mantel test of this metric according to the environmental distance derived from communtiy scores in the environmental PCA (two first axes). S|E is the partial Mantel test of this metric according to spatial distance conditionally to environment variation. The number of asterisk indicates significance threshold ( * p-value<0.05, ** p-value <0.01, *** p-value<0.001)._

```{r Ust & TAUst}
# Observed Hardys
spafun <- mapply(function(s, Ls){
  lapply(as.list(1:length(s)), function(i){
    if (length(which(is.na(s[i]))) > 0) {
      sel <- intersect(names(Ls$abundance), row.names(s[-which(is.na(s[i])),]))
      return(spacodi.calc(sp.plot = t(Ls$abundance[sel]), sp.traits = data.frame(s[sel,i], row.names = sel), pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
    } else {
      return(spacodi.calc(sp.plot = t(Ls$abundance), sp.traits = s, pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
    }
  })
}, s = Q, Ls = L, SIMPLIFY = F)
spafun <- lapply(spafun, function(s){
  names(s) <- names(Q$C)
  return(s)
})
## Mantel values
spafun.mantel <- mapply(function(s, Mes, Mss){
  lapply(s, function(t){
    list(
      `E` = lapply(t, function(M){
        vegan::mantel(M, Mes, permutations = 0)$statistic
      }), 
      `S | E` = lapply(t, function(M){
        vegan::mantel.partial(M, Mss, Mes, permutations = 0)$statistic
      })
    )
  })
}, s = spafun, Mes = Me, Mss = Ms, SIMPLIFY = F)

# Simulated mantel
cl <- makeCluster(cores)
clusterExport(cl, list('nQ', 'L', 'Me', 'Ms', 'spacodi.calc'))
nspafun.mantel <- clusterMap(cl, function(nQs, Ls, Mes, Mss){ 
  lapply(nQs, function(q){
    s <- lapply(as.list(1:length(q)), function(i){
      if (length(which(is.na(q[i]))) > 0) {
        sel <- intersect(names(Ls$abundance), row.names(q[-which(is.na(q[i])),]))
        return(spacodi.calc(sp.plot = t(Ls$abundance[sel]), sp.traits = data.frame(q[sel,i], row.names = sel), pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
      } else {
        return(spacodi.calc(sp.plot = t(Ls$abundance), sp.traits = q, pairwise = T)[c('pairwise.Ust', 'pairwise.TAUst')])
      }
    })
    names(s) <- names(q)
    res <- list(
      `E` = lapply(s, function(t){
        lapply(t, function(M){
          try(vegan::mantel(M, Mes, permutations = 0)$statistic)
        })
      }),
      `S | E` = lapply(s, function(t){
        lapply(t, function(M){
          try(vegan::mantel.partial(M, Mss, Mes, permutations = 0)$statistic)
        })
      })
    )
    res <- lapply(res, function(r){names(r) <- names(q) ; return(r)})
    res <- lapply(res, `[`, names(res[[1]]))
    res <- apply(do.call(rbind, res), 2, as.list)
    return(res)
  })
}, nQs = nQ, Ls = L, Mes = Me, Mss = Ms, SIMPLIFY = F)
stopCluster(cl)
rm(cl, nQ) ; invisible(gc())

## Non null mantels
nspafun.mantel <- lapply(nspafun.mantel, function(s){
  lapply(s, `[`, names(s[[1]]))
  apply(do.call(rbind, s), 2, as.list)
})
nspafun.mantel <- lapply(nspafun.mantel, function(s){
  lapply(s, function(m){
    lapply(m, `[`, names(m[[1]]))
    apply(do.call(rbind, m), 2, as.list)
  })
})
nspafun.mantel <- lapply(nspafun.mantel, function(s){
  lapply(s, function(m){
    lapply(m, function(u){
      lapply(u, `[`, names(u[[1]]))
      apply(do.call(rbind, u), 2, as.list)
    })
  })
})
nspafun.mantel <- lapply(nspafun.mantel, function(s){
  lapply(s, function(m){
    lapply(m, function(u){
      lapply(u, unlist)
    })
  })
})

spafun.res <- mapply(function(s, ns){
  mapply(function(t, nt){
    mapply(function(m, nm){
      mapply(function(u, nu){
        paste(round(u, 3), Uppangala::stars(min(rank(c(u,nu))[1]/(n+1), 1-rank(c(u,nu))[1]/(n+1)), ns = '   '))
      }, u = m, nu = nm)
    }, m = t, nm = nt)
  }, t = s, nt = ns, SIMPLIFY = F)
}, s = spafun.mantel, ns = nspafun.mantel, SIMPLIFY = F)
trait <- names(spafun.res$C)
spafun.res <- lapply(spafun.res, function(s){
  mapply(function(t, tr){
    t <- data.frame(t)
    t <- cbind(c(tr, ' '), c('$U_{st}$', '$\\tau _{st}$'),  t)
    names(t) <- c('T', 'M', 'E', 'S|E')
    return(t)
  }, t = s, tr = trait, SIMPLIFY = F)
})
spafun.res <-lapply(spafun.res, function(s){
  s <- do.call('rbind', s)
})
spafun.res <- mapply(function(s, str){
  s <- cbind(c(str, rep(' ', dim(s)[1]-1)), s)
  names(s)[1] <- 'S'
  return(s)
}, s = spafun.res, str = c('C', 'U'), SIMPLIFY = F)
spafun.res <- do.call('rbind', spafun.res)

# Result
kable(bold(spafun.res), row.names = F)
rm(spafun.mantel, nspafun.mantel, spafun.res)
```

@@

## S2: Phylogenetic varitions among communities

To investigate phylogenetic variation among communities we calculated $\beta _{st}$ and $\pi _{st}$, for phylogenetic composition in abundance and presence-absence respectively [@hardy_characterizing_2007]. $\beta _{st}$ integrate only phylogenetic turnover, weighting species by their local abundances; whereas $\pi _{st}$ express phylogenetic turnover, without accounting for species local abundances.

We used null models to test if observed correlation with Mantel statistic differed significantly from expectations under null hypothesis. $\beta st$ and $\pi st$ were tested against the nul model *"1s"* suggested by @hardy_testing_2008 which randomizes phylogenetic tree tips.

```{r BETAst $ PIst}
# Observed Hardys
phy <- tree
phy$tip.label <- Species$SpCode[match(gsub('_', ' ',phy$tip.label), Species$LatinName)]
tips <- lapply(L, function(s){
  c(setdiff(phy$tip.label, names(s$abundance)), setdiff(names(s$abundance), phy$tip.label))
})
phy <- lapply(tips, function(s){
  drop.tip(phy, s)
})
phy <- lapply(phy, function(s){
  s$node.label <- NULL
  return(s)
})
spaphy <- mapply(function(Ls, phys){
  spacodi.calc(sp.plot = t(Ls$abundance), phy = phys, pairwise = T)[c('pairwise.Bst', 'pairwise.PIst')]
}, Ls = L, phys = phy, SIMPLIFY = F)
## Mantel values
spaphy.mantel <- mapply(function(s, Mes, Mss){
  list(
    `E` = lapply(s, function(m){
      vegan::mantel(m, Mes, permutations = 0)$statistic
      }), 
    `S | E`= lapply(s, function(m){
      vegan::mantel.partial(m, Mss, Mes, permutations = 0)$statistic
      })
    )
}, s = spaphy, Mes = Me, Mss = Ms, SIMPLIFY = F)

# Simulated mantel
nphy <- lapply(phy, function(s){
  rep(list(s), 999)
})
cl <- makeCluster(cores)
clusterExport(cl, list('nphy', 'L', 'spacodi.calc', 'Ntip', 'resamp.phy', 'branching.times'))
nphy <- clusterApply(cl, nphy, function(s){
  lapply(s, resamp.phy)
})
stopCluster(cl)
nspaphy <- mapply(function(s, Ls){
  lapply(s, function(p){
  spacodi.calc(sp.plot = t(Ls$abundance), phy = p, pairwise = T)[c('pairwise.Bst', 'pairwise.PIst')]
  })
}, s = nphy, Ls = L, SIMPLIFY = F)
rm(cl, nphy)
invisible(gc())
nspaphy.mantel <- mapply(function(s, Mes, Mss){
  list(
    `E` =lapply(s, function(x){
      lapply(x, function(M){
        vegan::mantel(M, Mes, permutations = 0)$statistic})}), 
    `S | E` = lapply(s, function(x){
      lapply(x, function(M){
        vegan::mantel.partial(M, Mss, Mes, permutations = 0)$statistic
      })
    })
  )
}, s = nspaphy, Mes = Me, Mss = Ms, SIMPLIFY = F)
## Non null mantels
names(nspaphy.mantel) <- c('C', 'U')
nspaphy.mantel <- lapply(nspaphy.mantel, function(s){
  lapply(s, function(m){
    lapply(m, `[`, names(m[[1]]))
    apply(do.call(rbind, m), 2, as.list)
  })
})
nspaphy.mantel <- lapply(nspaphy.mantel, function(s){
  lapply(s, function(m){
    lapply(m, unlist)
  })
})
spaphy.res <- mapply(function(s, ns){
  mapply(function(m, nm){
    mapply(function(b, nb){
      paste(format(b, scientific = T, digits = 3), 
            Uppangala::stars(min(1-rank(c(b,nb))[1]/(n+1),rank(c(b,nb))[1]/(n+1)), ns = '   '))
    }, b = m, nb = nm, SIMPLIFY = T)
  }, m = s, nm = ns, SIMPLIFY = F)
}, s = spaphy.mantel, ns = nspaphy.mantel, SIMPLIFY = F)
spaphy.res <- mapply(function(s, str){
  s <- data.frame(s)
  s <- cbind(c(str, ' '), c('$U_{st}$', '$\\tau _{st}$'), s)
  names(s) <- c('S', 'M', 'E', 'S|E')
  return(s)
}, s = spaphy.res, str = c('C', 'U'), SIMPLIFY = F)
spaphy.res <- do.call('rbind', spaphy.res)

# Result
kable(bold(spaphy.res), row.names = F)
rm(mtest, spaphy.mantel, nspaphy, nspaphy.mantel, spaphy.res)
```
